---
title: "Bayesian Practice — Cleaned Pipeline"
author: "Theo Snow"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: united
params:
  input_csv: ""  
  seed: 123                                   
  bayes_draws: 1000                         
  treatment_prob: 0.5                       
  p0: 0.3                                     
  p1: 0.5                                     
---
Please Note: This github repo is solely a record of work. This is not meant as a reproducible pipeline, as the data leveraged for this project contains PHI. I have removed the data calls so this HTML knit displays errors after every cell. This is not representative of the quality of my code, but rather that nothing below will run without the data I have intentionally removed. 

## **0. Options and Helper Functions**
```{r}
# ---- options ----
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  error = TRUE
)
```

```{r setup, include=FALSE}

options(width = 120, scipen = 999)
set.seed(params$seed)

required_pkgs <- c(
  "tidyverse", "broom", "scales", "rlang", "glue", "patchwork"
)
invisible(lapply(required_pkgs, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(glue::glue("Required package '{pkg}' is not installed."))
  }
  library(pkg, character.only = TRUE)
}))

safe_select <- function(df, cols) {
  df %>% select(any_of(cols))
}

safe_load_csv <- function(path) {
  if (!file.exists(path)) {
    warning(glue::glue("File '{path}' not found. Falling back to synthetic toy data."))
    return(NULL)
  }
  read.csv(path)
}

simulate_treatment_and_outcomes <- function(df, treatment_prob, p0, p1, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  df %>%
    mutate(
      treatment = rbinom(nrow(.), size = 1, prob = treatment_prob),  # binary per row/stratum
      Y0 = rbinom(nrow(.), size = Sample_Strata_Denom, prob = p0),
      Y1 = rbinom(nrow(.), size = Sample_Strata_Denom, prob = p1),
      outcome = ifelse(treatment == 1, Y1, Y0),
      outcome_rate = outcome / Sample_Strata_Denom
    )
}

compute_deterministic_ate <- function(df, weight_col = "weights") {
  w <- sym(weight_col)
  df %>%
    summarise(
      ate = sum((!!w) * outcome * treatment) / sum((!!w) * treatment) -
            sum((!!w) * outcome * (1 - treatment)) / sum((!!w) * (1 - treatment))
    )
}

bayesian_ate_distribution <- function(df, draws = 1000) {
  replicate(draws, {
    weight_draws <- with(df, {
      rbeta(nrow(df), Posterior_Alpha, Posterior_Beta) * Sample_Total_Denom / Sample_Strata_Denom
    })
    sum(weight_draws * df$outcome * df$treatment) / sum(weight_draws * df$treatment) -
      sum(weight_draws * df$outcome * (1 - df$treatment)) / sum(weight_draws * (1 - df$treatment))
  })
}

fit_quasi_poisson <- function(df, formula, offset_col, weight_col = NULL) {
  fm <- formula
  offset_term <- rlang::expr(offset(log(!!sym(offset_col))))
  if (!is.null(weight_col)) {
    model <- glm(
      eval(fm),
      data = df,
      family = quasipoisson(link = "log"),
      offset = eval(offset_term),
      weights = df[[weight_col]]
    )
  } else {
    model <- glm(
      eval(fm),
      data = df,
      family = quasipoisson(link = "log"),
      offset = eval(offset_term)
    )
  }
  tidy_res <- broom::tidy(model, exponentiate = TRUE, conf.int = TRUE)
  tidy_res
}

ci_to_sd <- function(mean_val, ci_lower, ci_upper, z = 1.96) {
  (ci_upper - ci_lower) / (2 * z)
}
```

## **1. Load and Transform Data**
```{r}

merged_data_1 <- read.csv("")

set.seed(123)  # reproducible

merged_data_1_extended <- merged_data_1 %>%
  mutate(
    n = Sample_Strata_Denom,
    n_treated = rbinom(n(), size = n, prob = 0.5),
    n_control = n - n_treated,

    Y_treated = rbinom(n(), size = n_treated, prob = 0.5), 
    Y_control = rbinom(n(), size = n_control, prob = 0.3),  

    outcome = Y_treated + Y_control,

    treatment_prop = if_else(n > 0, n_treated / n, NA_real_),
    outcome_rate = if_else(n > 0, outcome / n, NA_real_),

    ate_stratum = case_when(
      n_treated > 0 & n_control > 0 ~ (Y_treated / n_treated) - (Y_control / n_control),
      TRUE ~ NA_real_
    )
  )
```

## **2. Simulate treatment/outcomes**
```{r}
required_cols <- c("Sample_Strata_Denom")
stopifnot(all(required_cols %in% names(merged_data_1)))

merged_data_1_extended <- simulate_treatment_and_outcomes(
  merged_data_1,
  treatment_prob = params$treatment_prob,
  p0 = params$p0,
  p1 = params$p1,
  seed = params$seed + 1
)

merged_data_1_extended %>% slice_head(n = 5)

```
## **3. ATE Estimation**
```{r}
stopifnot(c("weights", "Posterior_Alpha", "Posterior_Beta",
            "Sample_Total_Denom", "Sample_Strata_Denom", "treatment", "outcome") %in% names(merged_data_1_extended))

# Deterministic ATE
deterministic_ate <- compute_deterministic_ate(merged_data_1_extended, weight_col = "weights") %>%
  mutate(method = "Deterministic")

# Bayesian ATE draws
bayes_results <- bayesian_ate_distribution(merged_data_1_extended, draws = params$bayes_draws)
bayesian_summary <- tibble(
  ate_mean = mean(bayes_results),
  ate_lower = quantile(bayes_results, 0.025),
  ate_upper = quantile(bayes_results, 0.975),
  method = "Bayesian"
)

# Combine for plotting
comparison_df <- bind_rows(
  deterministic_ate %>%
    transmute(method, ate, lower = ate, upper = ate),
  bayesian_summary %>%
    transmute(method, ate = ate_mean, lower = ate_lower, upper = ate_upper)
)

comparison_df
```
## **4. ATE Visualization**
```{r}
ggplot(comparison_df, aes(x = method, y = ate, color = method)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, size = 1) +
  geom_hline(yintercept = 140, linetype = "dashed", color = "gray50") +
  labs(
    title = "Estimated ATE: Deterministic vs Bayesian Weights",
    subtitle = glue::glue("True ATE = {140} (dashed line)"),
    y = "ATE",
    x = NULL
  ) +
  scale_color_manual(values = c("Deterministic" = "#d95f02", "Bayesian" = "#1b9e77")) +
  theme_minimal() +
  theme(legend.position = "none")
```

## **5. Weight Distributions**
```{r}
bayes_weight_dist <- ggplot(merged_data_1, aes(x = weight_mean)) +
  geom_histogram(bins = 30, fill = "skyblue", alpha = 0.6) +
  labs(title = "Distribution of Bayesian Posterior Mean Weights", x = "Bayesian Mean Weight", y = "Count") +
  theme_minimal()

# Deterministic weight distribution
det_weight_dist <- ggplot(merged_data_1, aes(x = weights)) +
  geom_histogram(bins = 30, fill = "coral", alpha = 0.6) +
  labs(title = "Distribution of Deterministic Weights", x = "Deterministic Weight", y = "Count") +
  theme_minimal()

# Side-by-side
bayes_weight_dist + det_weight_dist + plot_annotation(title = "Weight Comparisons")
```
## **6. Stochastic Weight Model**
```{r}
stopifnot(all(c("Year", "Sample_Strata_Denom", "Sample_Total_Denom") %in% names(merged_data_1)))

# Deterministic model: weight_mean treated as deterministic weight (example)
mod_det <- fit_quasi_poisson(
  df = merged_data_1,
  formula = quote(Sample_Strata_Denom ~ as.factor(Year)),
  offset_col = "Sample_Total_Denom",
  weight_col = "weight_mean"
) %>%
  mutate(model = "Deterministic") %>%
  mutate(mean_estimate = estimate)

simulate_yearly_model <- function(df, n_sim = 10) {
  coef_list <- vector("list", n_sim)
  for (i in seq_len(n_sim)) {
    df_sim <- df %>%
      mutate(
        weight_sim = rbeta(n(), Posterior_Alpha, Posterior_Beta)
      ) %>%
      mutate(
        weight_sim = weight_sim * (sum(weight_mean, na.rm = TRUE) / sum(weight_sim, na.rm = TRUE))  # <- normalization example
      )
    tmp <- fit_quasi_poisson(
      df = df_sim,
      formula = quote(Sample_Strata_Denom ~ as.factor(Year)),
      offset_col = "Sample_Total_Denom",
      weight_col = "weight_sim"
    ) %>%
      mutate(sim_id = i)
    coef_list[[i]] <- tmp
  }
  bind_rows(coef_list)
}
coef_stochastic_raw <- simulate_yearly_model(merged_data_1, n_sim = 10)
```
## **7. Stochastic Model Output**
```{r}
stoch_summary <- coef_stochastic_raw %>%
  filter(str_detect(term, "^as.factor\\(Year\\)")) %>%
  mutate(year = str_remove(term, "as.factor\\(Year\\)")) %>%
  group_by(year) %>%
  summarise(
    mean_estimate = mean(estimate),
    lower_95 = quantile(estimate, 0.025),
    upper_95 = quantile(estimate, 0.975),
    .groups = "drop"
  ) %>%
  mutate(model = "Stochastic")

det_summary <- mod_det %>%
  filter(str_detect(term, "^as.factor\\(Year\\)")) %>%
  mutate(
    year = str_remove(term, "as.factor\\(Year\\)"),
    mean_estimate = estimate,
    lower_95 = conf.low,
    upper_95 = conf.high,
    model = "Deterministic"
  ) %>%
  select(year, mean_estimate, lower_95, upper_95, model)

combined_model_estimates <- bind_rows(stoch_summary, det_summary) %>%
  mutate(year = factor(year, levels = sort(unique(year))))
```

##**8. Stochastic Model Visualization**
```{r}
ggplot(combined_model_estimates, aes(x = year, y = mean_estimate, color = model, group = model)) +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  geom_line(position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(ymin = lower_95, ymax = upper_95), width = 0.2,
                position = position_dodge(width = 0.4), alpha = 0.7) +
  labs(
    title = "Yearly Coefficient Estimates: Deterministic vs Stochastic",
    x = "Year",
    y = "Exponentiated Coefficient (Rate Ratio)",
    color = "Model"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Deterministic" = "#d95f02", "Stochastic" = "#1b9e77")) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40")
```



